Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-21 submission/manuscript.Rmd
results edit 3

Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. A subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A, top)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020]. While some germline-enriched genes are also expressed in na&#239;ve embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. Therefore, we tested if KDM5C was necessary for silencing germline genes at this developmental stage by evaluating the impact of _Kdm5c_ loss in male EpiLCs.

We first identified _Kdm5c_-KO EpiLC DEGs in our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and then assessed tissue-enriched gene expression. Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B). Markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_, were unaltered in _Kdm5c_-KO EpiLCs (Figure 3C). Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3D), indicating KDM5C loss does not impair EpiLC formation.

We next compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline DEGs are constitutively dysregulated or change over the course of development. We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F). EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023]. However, we did not see misexpression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell-like state (Figure 3H). Altogether, these results demonstrate that while the _Kdm5c_-KO brain primarily expresses spermiogenesis genes, _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis.

>>>>

Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. A subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020]. While some germline-enriched genes are also expressed in na&#239;ve embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. Therefore, we tested if KDM5C was necessary for the initial silencing germline genes in somatic lineages by evaluating the impact of _Kdm5c_ loss in male EpiLCs.

We first identified _Kdm5c_-KO EpiLC DEGs in our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and then assessed tissue-enriched gene expression. Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B). Markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_, were unaltered in _Kdm5c_-KO EpiLCs (Figure 3C). Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3D), indicating the misexpression of tissue-enriched genes with KDM5C loss does not impair EpiLC formation.

We next compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development. We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F). EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023]. However, we did not see misexpression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell-like state. Altogether, these results demonstrate that while the _Kdm5c_-KO brain primarily expresses spermiogenesis genes, _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis.
